Synaptic connectome of the ctenophore apical organ

### 2025-02-13 manuscript/Manuscript.qmd
MS: update refs

Recent comparative genomic studies (Ryan et al., 2013; Whelan et al., 2015; Whelan et al., 2017; Li et al., 2021; Whelan and Halanych, 2023) and chromosome-level synteny analyses across different animal phyla (Schultz et al., 2023) suggest that ctenophores are the sister group to all other metazoans and were the first animals to evolve a nervous system. This has led to attempts to compare genes and molecular modules in ctenophores using single-cell analyses, in which neuropeptides have been employed as neuronal markers (Sebé-Pedrós et al., 2018; Sachkova et al., 2021; Hayakawa et al., 2022). However, the homologous relationships between ctenophore genes and those of other animals are highly divergent, limiting the feasibility of sequence-based comparisons. Given these challenges, we considered an alternative approach: linking neural circuits to physiological responses to gain insight into neural function.

@@

Among these, we focused on the apical organ of ctenophores. In many bilaterian and non-bilaterian animals, an apical nervous system is located at one pole of the body, forming a specialized sensory structure that governs physiological functions and motor activities. In ctenophores, ultrastructural analyses have revealed that the aboral pole contains a diverse array of sensory cells and neuronal components (Hernandez-Nicaise, 1968, 1973, 1974; Tamm, 1982). Additionally, immunostaining using an anti-tyrosylated α-tubulin antibody has identified a neural structure called the deep nerve net within the apical organ, suggesting its involvement in neural regulation of sensory-motor control (Jager et al., 2011).

In this study, we specifically examined the gravity-sensing system within the apical organ. The apical organ of ctenophores contains a dense mineralized mass known as the statocyst, supported by four mechanosensory ciliated cells called balancer cells [@Noda_2014]. The ciliary movements of these balancer cells function as pacemakers for the eight rows of comb plates that control posture and swimming through mechanotransduction [@Tamm_2014] (Tamm, 1982). Synaptic connections projecting to the balancer cells have also been reported [@aronova1974electron]. Therefore, we conducted a connectome analysis to characterize the network structure of the apical organ and elucidate the neural circuits that link sensory input to behavioral responses in ctenophores.

@@

First, we embedded the entire body of the *Mnemiopsis leidyi* cydippid-stage larvae, five days post-fertilization, in Epon resin following high-pressure freezing. To reconstruct all synaptic-resolution connections among the cells comprising the apical organ, also known as the aboral organ, we prepared approximately 1,000 ultra-thin serial sections (50 nm thick) from the aboral tip of the larval body embedded in the resin block. Using a scanning electron microscope (Gemini SEM 500) with a resolution of 2.0 nm/pixel, we imaged only the region containing the apical organ. Subsequently, 620 of these images were stitched and aligned using TrakEM2. The resulting volumetric EM dataset had dimensions of 60 μm × 40 μm × 30 μm. We traced and annotated all cells in this dataset, ultimately reconstructing 909 cells, each with both nuclei and cell bodies intact.

@@

Previous electron microscopy studies have suggested the presence of various sensory cells, including photoreceptor and mechanoreceptor cells, within the aboral organ of ctenophores (Horridge, 1965; Krisch, 1973; Aronova, 1974; Vinnikov, 1974). By comparing our dataset with previously reported ultrastructural data, we successfully identified several sensory cells within the apical organ. However, notably, we did not find any clear presynaptic structures between these sensory cells and either neural or other cellular targets. In prior studies, immunostaining for specific neuropeptides in ctenophores has revealed positive signals in distinct cell populations within the apical organ (Sachkova et al., 2021; Hayakawa et al., 2022). These findings suggest that many sensory cells in the apical organ may primarily rely on neuropeptide-mediated paracrine signaling rather than direct synaptic connections, supporting the chemical brain hypothesis (Jekely, 2021). While paracrine transmission is generally slower than synaptic transmission, it allows for a greater diversity of information processing.

Conversely, previous research has demonstrated that mechanoreceptive sensory cells equipped with cilia (referred to as type 3 and type 4 cells) in ctenophores form direct synaptic connections with the sub-epithelial nerve net (SNN) and mesogleal neurons (Burkhurdt et al., 2023). Based on their ultrastructural features, these sensory cells are believed to respond to physical stimuli such as water flow, vibrations, and direct contact, transmitting information rapidly to cilia and muscle cells (Hertwig, 1880; Chun, 1880; Eimer, 1880).

Surprisingly, our study identified bridge cells as the primary cell type forming synaptic inputs onto neurons in the apical organ. Bridge cells were first described by Tamm et al. (2002) based on their morphological characteristics, with cellular projections extending to the basal region of balancer cilia [@Tamm_2002] (Tamm & Tamm, 2002). Our findings suggest that bridge cells are electrically coupled to balancer cilia and provide rapid feedback to neurons via synaptic transmission. Future studies analyzing differential responses to various sensory stimuli will not only elucidate the functional roles of bridge cells but also provide insights into the evolutionary relationship between transmission speed and synapse development.

@@

Our study suggests that differences in neural circuits contribute to variations in the timing of ciliary arrest, which may play a role in modulating swimming direction in ctenophores (Richard Satterlie, 2015). This finding provides evidence that increased circuit complexity contributes to behavioral diversification. By integrating multiple neurons and altering circuit structure, information can be branched, allowing for diverse motor patterns within the framework of fast synaptic transmission.

@@

From a broader neuroanatomical perspective, nervous system structures across various animals have been shaped through selective optimization to adapt to environmental constraints (Bullmore & Sporns, 2012; Perin et al., 2011). For example, in nematodes, the spatial arrangement of ganglia is optimized to minimize the total length of neural wiring (White et al., 1986). Similarly, in ctenophores, understanding how transmission via syncytial neurons and synaptic connections has been selectively optimized at the circuit level will provide crucial insights into the evolutionary scaling of neural networks from early nervous systems to complex brains (Farnworth and Montgomery, 2024). In particular, this research may help elucidate how circuit modularization and layered structures emerged over evolutionary time.

@@

The following EM image stacks, including all traces and annotations, are available at https://catmaid.jekelylab.ex.ac.uk. These datasets can be queried either within CATMAID or through the CATMAID application programming interface (API) using the catmaid or natverse packages (Bates et al., 2020). The dataset encompasses all EM images (in JPG format), skeletons, meshes, node tags, connectors, and annotations. Additionally, we provide all R scripts used for data acquisition and figure generation (Jokura et al., 2025). All plots, figures (including anatomical renderings), and figure layouts can be fully reproduced using the provided R scripts. While the scripts are mostly organized by figure, some general-purpose scripts—for tasks such as loading libraries, accessing CATMAID data, and defining common functions—are shared across multiple figures.

@@

The reconstruction of cells and connectivity was performed using CATMAID (Saalfeld et al., 2009), analyzing a total of 620 serial sections. To digitally reconstruct all neurons within the serial TEM dataset, we utilized the collaborative web-based application CATMAID, installed on a local server (Saalfeld et al., 2009; Schneider-Mizell et al., 2016). To mark the locations of cell bodies, we placed tags at the approximate center of each nucleus within the dataset. At each nuclear center, the radius of the single node was adjusted according to the size of the cell body in that specific layer. All skeletons were rooted at their respective cell bodies, and nodes were tagged as "soma." Synapses were identified based on four key structural features: the cell membrane, synaptic vesicles, the endoplasmic reticulum, and mitochondria. Most synapses could be verified across consecutive sections, ensuring accurate annotation and connectivity mapping.

@@

We conducted the analysis using the cydippid stage of Mnemiopsis leidyi at five days post-fertilization. A coverslip with a thin layer of Vaseline applied to two edges was lightly placed on a slide glass. Filtered natural seawater was introduced under the coverslip, along with the cydippids. By gently moving the coverslip, the desired orientation for observation was adjusted, and once positioned, slight pressure was applied to immobilize the cydippids. To stabilize the statolith's position, the microscope was tilted 90 degrees, arranging the stage vertically. The movement of the balancer cilia was observed using a differential interference contrast (DIC) microscope (Zeiss Axio Imager.M2) equipped with a highly sensitive CMOS monochrome camera optimized for the near-infrared (NIR) range (UI-3360CP-NIR-GL Rev.2, iDS). Observations were made using a 40x objective lens (Objective LD LCI Plan-Apochromat 40x/1.2 Imm Corr DIC M27) with glycerine immersion. For recording, we employed a custom-made NIR LED strobe illumination system (wavelength: 850 nm) synchronized with the camera's exposure signals. The camera and LED strobe system were operated using the Video Capture Software BohNavi. Images were recorded at a resolution of 640×480 pixels with a frame rate of 100 fps for 2 minutes, using 0.05 ms pulses from the LED. To analyze the ciliary beating of the balancer cilia, we utilized the Multi Kymograph project in Fiji software.
=======
Figure 6 - cilia? renderings of basal bodies, centrioles, length, axoneme structure (table), table about cell types, number of cilia etc.

@@

"synaptic regions with the characteristic presynaptic triad mor- phology of ctenophore nerves (i.e., a single layer of vesicles, smooth ER sac, and closely apposed mito- chondria [@Horridge_1964; @Hernandez_Nicaise_1973] Hernandez- Nicaise, 1973)"

@@

>>>>>>> Stashed changes

@@

```{=html}
<!--
## See https://github.com/hypothesis/client
## Add commenting functionality with this script
-->
```
```{=html}
<script src="hypothes.is/embed.js" async></script>
```
\newpage

>>>>

Recent comparative genomic studies [@Ryan2013; @Whelan2015a; @Whelan2017; @Whelan2023; @Li2021] and chromosome-level synteny analyses across different animal phyla [@Schultz2023] suggest that ctenophores are the sister group to all other metazoans and were the first animals to evolve a nervous system. This has led to attempts to compare genes and molecular modules in ctenophores using single-cell analyses, in which neuropeptides have been employed as neuronal markers [@Sebé-Pedrós2018; @Sachkova2021; @Hayakawa2022]. However, the homologous relationships between ctenophore genes and those of other animals are highly divergent, limiting the feasibility of sequence-based comparisons. Given these challenges, we considered an alternative approach: linking neural circuits to physiological responses to gain insight into neural function.

@@

Among these, we focused on the apical organ of ctenophores. In many bilaterian and non-bilaterian animals, an apical nervous system is located at one pole of the body, forming a specialized sensory structure that governs physiological functions and motor activities. In ctenophores, ultrastructural analyses have revealed that the aboral pole contains a diverse array of sensory cells and neuronal components [@Hernandez-Nicaise1968; @Hernandez_Nicaise_1973; @Hernandez-Nicaise1974] <!--# (Tamm, 1982 ?) -->. Additionally, immunostaining using an anti-tyrosylated α-tubulin antibody has identified a neural structure called the deep nerve net within the apical organ, suggesting its involvement in neural regulation of sensory-motor control [@Jager2011].

In this study, we specifically examined the gravity-sensing system within the apical organ. The apical organ of ctenophores contains a dense mineralized mass known as the statocyst, supported by four mechanosensory ciliated cells called balancer cells [@Noda_2014]. The ciliary movements of these balancer cells function as pacemakers for the eight rows of comb plates that control posture and swimming through mechanotransduction [@Tamm_2014] <!--# (Tamm, 1982 ?) -->. Synaptic connections projecting to the balancer cells have also been reported [@aronova1974electron]. Therefore, we conducted a connectome analysis to characterize the network structure of the apical organ and elucidate the neural circuits that link sensory input to behavioral responses in ctenophores.

@@

First, we embedded the entire body of the *Mnemiopsis leidyi* cydippid-stage larvae, five days post-fertilization, in Epon resin following high-pressure freezing. To reconstruct all synaptic-resolution connections among the cells comprising the apical organ, also known as the aboral organ, we prepared approximately 1,000 ultra-thin serial sections (50 nm thick) from the aboral tip of the larval body embedded in the resin block. Using a scanning electron microscope (Zeiss Gemini 500) with a resolution of 2.0 nm/pixel, we imaged only the region containing the apical organ. Subsequently, 620 of these images were stitched and aligned using TrakEM2. The resulting volumetric EM dataset had dimensions of 60 μm × 40 μm × 30 μm. We traced and annotated all cells in this dataset, ultimately reconstructing 909 cells, each with both nuclei and cell bodies intact.

@@

Previous electron microscopy studies have suggested the presence of various sensory cells, including photoreceptor and mechanoreceptor cells, within the aboral organ of ctenophores [@HORRIDGE1965; @vinnikov1974e] <!--#  Krisch, 1973; Aronova, 1974 ? --> ). By comparing our dataset with previously reported ultrastructural data, we successfully identified several sensory cells within the apical organ. However, notably, we did not find any clear presynaptic structures between these sensory cells and either neural or other cellular targets. In prior studies, immunostaining for specific neuropeptides in ctenophores has revealed positive signals in distinct cell populations within the apical organ [@Sachkova2021; @Hayakawa2022]. These findings suggest that many sensory cells in the apical organ may primarily rely on neuropeptide-mediated paracrine signaling rather than direct synaptic connections, supporting the chemical brain hypothesis (Jekely, 2021). While paracrine transmission is generally slower than synaptic transmission, it allows for a greater diversity of information processing.

Conversely, previous research has demonstrated that mechanoreceptive sensory cells equipped with cilia (referred to as type 3 and type 4 cells) in ctenophores form direct synaptic connections with the sub-epithelial nerve net (SNN) and mesogleal neurons [@Burkhardt2023]. Based on their ultrastructural features, these sensory cells are believed to respond to physical stimuli such as water flow, vibrations, and direct contact, transmitting information rapidly to cilia and muscle cells <!--# add these to bib file -->(Hertwig, 1880; Chun, 1880; Eimer, 1880).

Surprisingly, our study identified bridge cells as the primary cell type forming synaptic inputs onto neurons in the apical organ. Bridge cells were first described by Tamm et al. based on their morphological characteristics, with cellular projections extending to the basal region of balancer cilia [@Tamm_2002]. Our findings suggest that bridge cells are electrically coupled to balancer cilia and provide rapid feedback to neurons via synaptic transmission. Future studies analyzing differential responses to various sensory stimuli will not only elucidate the functional roles of bridge cells but also provide insights into the evolutionary relationship between transmission speed and synapse development.

@@

Our study suggests that differences in neural circuits contribute to variations in the timing of ciliary arrest, which may play a role in modulating swimming direction in ctenophores [@Satterlie2015a]. This finding provides evidence that increased circuit complexity contributes to behavioral diversification. By integrating multiple neurons and altering circuit structure, information can be branched, allowing for diverse motor patterns within the framework of fast synaptic transmission.

@@

From a broader neuroanatomical perspective, nervous system structures across various animals have been shaped through selective optimization to adapt to environmental constraints (Bullmore & Sporns, 2012; Perin et al., 2011). For example, in nematodes, the spatial arrangement of ganglia is optimized to minimize the total length of neural wiring (White et al., 1986). Similarly, in ctenophores, understanding how transmission via syncytial neurons and synaptic connections has been selectively optimized at the circuit level will provide crucial insights into the evolutionary scaling of neural networks from early nervous systems to complex brains [@Farnworth2024]. In particular, this research may help elucidate how circuit modularization and layered structures emerged over evolutionary time.

@@

The following EM image stacks, including all traces and annotations, are available at https://catmaid.jekelylab.ex.ac.uk. These datasets can be queried either within CATMAID or through the CATMAID application programming interface (API) using the catmaid or natverse packages [@Bates2020]. The dataset encompasses all EM images (in JPG format), skeletons, meshes, node tags, connectors, and annotations. Additionally, we provide all R scripts used for data acquisition and figure generation (Jokura et al., 2025). All plots, figures (including anatomical renderings), and figure layouts can be fully reproduced using the provided R scripts. While the scripts are mostly organized by figure, some general-purpose scripts—for tasks such as loading libraries, accessing CATMAID data, and defining common functions—are shared across multiple figures.

@@

The reconstruction of cells and connectivity was performed using CATMAID (Saalfeld et al., 2009), analyzing a total of 620 serial sections. To digitally reconstruct all neurons within the serial TEM dataset, we utilized the collaborative web-based application CATMAID, installed on a local server [@Saalfeld2009; @Schneider-Mizell2016]. To mark the locations of cell bodies, we placed tags at the approximate center of each nucleus within the dataset. At each nuclear center, the radius of the single node was adjusted according to the size of the cell body in that specific layer. All skeletons were rooted at their respective cell bodies, and nodes were tagged as "soma." Synapses were identified based on four key structural features: the cell membrane, synaptic vesicles, the endoplasmic reticulum, and mitochondria. Most synapses could be verified across consecutive sections, ensuring accurate annotation and connectivity mapping.

@@

We conducted the analysis using the cydippid stage of *Mnemiopsis leidyi* at five days post-fertilization. A coverslip with a thin layer of Vaseline applied to two edges was lightly placed on a slide glass. Filtered natural seawater was introduced under the coverslip, along with the cydippids. By gently moving the coverslip, the desired orientation for observation was adjusted, and once positioned, slight pressure was applied to immobilize the cydippids. To stabilize the statolith's position, the microscope was tilted 90 degrees, arranging the stage vertically. The movement of the balancer cilia was observed using a differential interference contrast (DIC) microscope (Zeiss Axio Imager.M2) equipped with a highly sensitive CMOS monochrome camera optimized for the near-infrared (NIR) range (UI-3360CP-NIR-GL Rev.2, iDS). Observations were made using a 40x objective lens (Objective LD LCI Plan-Apochromat 40x/1.2 Imm Corr DIC M27) with glycerine immersion. For recording, we employed a custom-made NIR LED strobe illumination system (wavelength: 850 nm) synchronized with the camera's exposure signals. The camera and LED strobe system were operated using the Video Capture Software BohNavi. Images were recorded at a resolution of 640×480 pixels with a frame rate of 100 fps for 2 minutes, using 0.05 ms pulses from the LED. To analyze the ciliary beating of the balancer cilia, we utilized the Multi Kymograph project in Fiji software.

Figure 6 - cilia? renderings of basal bodies, centrioles, length, axoneme structure (table), table about cell types, number of cilia etc.

@@

"synaptic regions with the characteristic presynaptic triad mor- phology of ctenophore nerves (i.e., a single layer of vesicles, smooth ER sac, and closely apposed mito- chondria [@Horridge_1964; @Hernandez_Nicaise_1973] "

@@

@@

```{=html}
<!--
## See https://github.com/hypothesis/client
## Add commenting functionality with this script
-->
```

```{=html}
<script src="hypothes.is/embed.js" async></script>
```

\newpage
